GeoSPARQL 1.1: an almost decadal update to the most important geospatial LOD standard

### 2021-03-15 manuscript.tex
Update on Overleaf.

\begin{abstract}
In 2012 the Open Geospatial Consortium published GeoSPARQL 
defining ``SPARQL extension functions'', ``RIF rules'', ``and RDF/OWL ontology for 
[spatial] information'' and supporting vocabularies.\\

@@

\section{Introduction}\label{sec:introduction}
The GeoSPARQL standard, issued in 2012 by the Open Geospatial Consortium (OGC)\footnote{\url{https://www.ogc.org}} 
is one of, if not the most\footnote{It's hard to calculate use but references to GeoSPARQL other well-known standards, 
such as DCAT2 (\url{https://www.w3.org/TR/vocab-dcat/}) suggests this}, used \textit{Sematic Web} standards for 
spatial data.

@@

Recently, 2019, the OGC reconstituted a \textit{GeoSPARQL Standards Working Group} (SWG) to update GeoSPARQL. The general 
motivation for work within the area of GeoSPARQL, that of \textit{Semantic Web} spatial data, and a series of
fault fixes and proposed extensions to GeoSPARQL 1.0 are captured in an OGC White Paper~\cite{geosparqlwhitepaper}. Some,
but not all, of the SDWWG's ideas have been taken up by the SDW, for example the \textit{Best Practices}~\cite{van_den_brink_best_2018}
aspiration that ``A possible way forward is an update for the GeoSPARQL spatial ontology. This would provide an 
agreed spatial ontology, i.e., a bridge or common ground between geographical and non-geographical spatial data...''.
This point has been considered by the SWG and included in future releases' scope.

@@

\begin{itemize}
    \item[$\ast$] new geometry serializations
    \begin{itemize}
        \item[$-$] GeoJSON, KML itc. Newer, popular format
    \end{itemize} 
    \item[$\ast$] new ontology classes to cater for more nuanced spatial information
    \item[$\ast$] more spatial functions
    \begin{itemize}
        \item[$-$] implementing functions well-known in non \texttt{Semantic Web} spatial systems
    \end{itemize} 
    \item[$\ast$] scalar spatial properties 
    \begin{itemize}
        \item[$-$] area, volume etc. alongside geometries
    \end{itemize} 
    \item[$\ast$] better handling of Spatial (Coordinate) Reference Systems (SRS)
    \begin{itemize}
        \item[$-$] potentially allowing for automated coordinate serialization conversions
    \end{itemize} 
    \item[$\ast$] Internet protocol-based selection of different geometries for features
\end{itemize}

@@

\section{Updates in GeoSPARQL 1.1}\label{sec:newfeatures}
So far (as of June, 2020) the GeoSPARQL SWG has triaged change requests for GeoSPARQL releases and has addressed
many 1.1 requests which are the only ons we report here. The next sections foreshadow likely 1.2 and 2.0 updates.

@@

The profile declaration for GeoSPARQL 1.0 will be published by the OGC as a stand-alone resource sometime in early 2021 along with some 
updated GeoSPARQL 1.0 resources. The profile declaration for GeoSPARQL 1.1 will be published at the same time as the 
1.1 releases, currently expected in mid-2021. All 1.0 and 1.1 relase resources are currently available in draft form 
in the SWG's online code repository\footnote{\url{https://github.com/opengeospatial/ogc-geosparql}}. The 1.1 releases' resources are:

@@

\subsection{New spatial functions}\label{sec:newfunctions}
While spatial aggregation functions are the norm in many non-semantic geospatial databases such as POSTGIS or Oracle Spatial, 
at the time of defining the GeoSPARQL 1.0 standard, aggregation functions had not yet been introduced into the SPARQL standard,
but have been since with SPARQL 1.1~\cite{w3c_sparql_working_group_sparql_2013}. Spatial aggregation functions 
similar to traditional (relation database), aggregation functions such as AVG, MAX, or MIN allow aggregated results of geometry 
queries, for example, to create the union of a set of selected serialized geometries. While calculating these aggregates is 
certainly possible outside of a semantic database, and thus GeoSPARQL, the inclusion of the functions provides distinct advantages:

@@

\begin{figure}[htb]
    \centering
    \includegraphics[width=\linewidth]{images/geold_ontology.png}
    \caption{GeoSPARQL 1.1 ontology including one example feature}
    \label{fig:geosparql11ontology}
\end{figure}

\section{Conclusions}\label{sec:conclusions}
A staged schedule of updates to this important \textit{Semantic Web} spatial standard has been initiated with simple and strictly backwards-compatable changes now in GeoSPARQL 1.1. 1.2 will include more detailed work later in 2021 and 2.0, as yet un-specified, might be somewhat different: to be determined by more requirements gathering and feedback on the 1.1 and 1.2 releases.

\subsection{Future Work}\label{sec:futurework}
\begin{itemize}
    \item[$\ast$] GeoSPARQL 1.1: to be completed by July, 2021
    \item[$\ast$] GeoSPARQL 1.2: more adventurous extension - late 2021
    \item[$\ast$] GeoSPARQL 2.0: so-far unspecified
\end{itemize}

>>>>

\begin{abstract}
In 2012 the Open Geospatial Consortium published GeoSPARQL 
defining ``SPARQL extension functions'', ``RIF rules'', ``an RDF/OWL ontology for [spatial] information'' and supporting vocabularies.\\

@@

\section{Introduction}\label{sec:introduction}
The GeoSPARQL standard, issued in 2012 by the Open Geospatial Consortium (OGC)\footnote{\url{https://www.ogc.org}} 
is one of, if not the most\footnote{It's hard to calculate use but references to GeoSPARQL other well-known standards, 
such as DCAT2 (\url{https://www.w3.org/TR/vocab-dcat/}) or CIDOC-CRM \cite{doerr2013crmgeo} suggests this}, used \textit{Semantic Web} standards for 
spatial data.

@@

Recently, in 2019, the OGC reconstituted a \textit{GeoSPARQL Standards Working Group} (SWG) to update GeoSPARQL. The general 
motivation for work within the area of GeoSPARQL, that of \textit{Semantic Web} spatial data, and a series of
fault fixes and proposed extensions to GeoSPARQL 1.0 are captured in an OGC White Paper~\cite{geosparqlwhitepaper}. Some,
but not all, of the SDWWG's ideas have been taken up by the SDW, for example the \textit{Best Practices}~\cite{van_den_brink_best_2018}
aspiration that ``A possible way forward is an update for the GeoSPARQL spatial ontology. This would provide an 
agreed spatial ontology, i.e., a bridge or common ground between geographical and non-geographical spatial data...''.
This point has been considered by the SWG and included in future releases' scope.

@@

\begin{itemize}
    \item[$\ast$] new geometry serializations
    \begin{itemize}
        \item[$-$] GeoJSON, KML etc. Newer, popular formats
    \end{itemize} 
    \item[$\ast$] new ontology classes to cater for more nuanced spatial information
    \item[$\ast$] more spatial functions
    \begin{itemize}
        \item[$-$] implementing functions well-known in non \texttt{Semantic Web} spatial systems
    \end{itemize} 
    \item[$\ast$] scalar spatial properties 
    \begin{itemize}
        \item[$-$] area, volume etc. alongside geometries
    \end{itemize} 
    \item[$\ast$] better handling of Spatial (Coordinate) Reference Systems (SRS)
    \begin{itemize}
        \item[$-$] potentially allowing for automated coordinate serialization conversions
    \end{itemize} 
    \item[$\ast$] Internet protocol-based selection of different geometries for features
\end{itemize}

@@

\section{Updates in GeoSPARQL 1.1}\label{sec:newfeatures}
So far (as of June, 2020) the GeoSPARQL SWG has triaged change requests for GeoSPARQL releases and has addressed
many 1.1 requests which are the only ones we report here. The next sections foreshadow likely 1.2 and 2.0 updates.

@@

The profile declaration for GeoSPARQL 1.0 will be published by the OGC as a stand-alone resource sometime in early 2021 along with some 
updated GeoSPARQL 1.0 resources. The profile declaration for GeoSPARQL 1.1 will be published at the same time as the 
1.1 releases, currently expected in mid-2021. All 1.0 and 1.1 release resources are currently available in draft form 
in the SWG's online code repository\footnote{\url{https://github.com/opengeospatial/ogc-geosparql}}. The 1.1 releases' resources are:

@@

\subsection{New spatial functions}\label{sec:newfunctions}
While spatial aggregation functions are the norm in many non-semantic geospatial databases such as PostGIS or Oracle Spatial, 
at the time of defining the GeoSPARQL 1.0 standard, aggregation functions had not yet been introduced into the SPARQL standard,
but have been with SPARQL 1.1~\cite{w3c_sparql_working_group_sparql_2013}. Spatial aggregation functions 
similar to traditional (relation database), aggregation functions such as AVG, MAX, or MIN allow aggregated results of geometry 
queries, for example, to create the union of a set of selected serialized geometries. While calculating these aggregates is 
certainly possible outside of a semantic database, and thus GeoSPARQL, the inclusion of the functions provides distinct advantages:

@@

\begin{figure}[htb]
    \centering
    \includegraphics[width=\linewidth]{images/geold_ontology.png}
    \caption{Excerpt of the GeoSPARQL 1.1 ontology including one example feature}
    \label{fig:geosparql11ontology}
\end{figure}

\section{Conclusions}\label{sec:conclusions}
A staged schedule of updates to this important \textit{Semantic Web} spatial standard has been initiated with simple and strictly backwards-compatible changes now in GeoSPARQL 1.1. Features discussed for GeoSPARQL 1.2 include the formalization of coordinate reference systems in RDF, the depiction of accurracies and level of detail and the addition of further - possibly also binary - literal types. Work on GeoSPARQL 1.2 will start later in 2021. GeoSPARQL 2.0, as yet un-specified, is likely to introduce more substantial changes to the standard. Changes proposed for GeoSPARQL 2.0 include to broaden the scope of GeoSPARQL to further kinds of spatial data. To that end, full-featured support for 3D geometries and support for coverages are discussed on the level of data representations. These proposals are related to some growing interest in the semantic web community in representing further geospatial data related to building modeling information \cite{zhang2018bimsparql} and coverage data \cite{homburg2020geosparql+}. More requirements might also be introduced once feedback has been received from the GeoSPARQL 1.1 and 1.2 releases.
